---
title: "Data Extraction: Extracting Raw Airline-Flight Level Data to Airline-Route Level Data"
author: "Eugene Kiepeeh"
date: "2026-09-02"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
```

# Run Full Data Extraction and Processing
Below, the raw data of each flight record for each month since 2020 and 2025 are extracted locally and processed into a tabular data were each row is 

- a route completed by an airline in a month and it tells you the number of trips assigned(`totalFlghts`)---inclusive of canceled and non-canceled, and the percentage of those trips delayed or canceled represented by `dep_delay_rate`, `arr_delay_rate`. 

```{r Loading Final Data}
process_month <- function(file_path) {
  
  cat("\n==============================\n")
  #cat("Processing file:", file_id, "\n")
  cat("Path:", file_path, "\n")
  
  df <- read_csv(file_path, show_col_types = FALSE)
  
  cat("Initial dimensions:", nrow(df), "rows |", ncol(df), "cols\n")
  
  df <- read_csv(file_path)
  
  clean <- df |> select(-any_of(c(
  "MKT_CARRIER_AIRLINE_ID", "MKT_CARRIER_FL_NUM",
  "ORIGIN_AIRPORT_ID", "ORIGIN_AIRPORT_SEQ_ID",
  "ORIGIN_CITY_MARKET_ID", "DEST_AIRPORT_ID",
  "DEST_AIRPORT_SEQ_ID", "DEST_CITY_MARKET_ID",
  "DISTANCE_GROUP", "FLIGHTS", "CANCELLATION_CODE")))

routeLevel_df <- clean |>
  group_by(YEAR, MONTH, BRANDED_CODE_SHARE, ORIGIN, DEST) |>
  mutate(across(c(DEP_TIME, ARR_TIME), as.numeric)) |>
  # ind. airline performance on a route
  summarize(totalFlights = n(), 
            n_DepDelay = sum(DEP_DEL15 == 1, na.rm = T),
            depDelay_rate = n_DepDelay/totalFlights,
            
            n_ArrDelay = sum(ARR_DEL15 == 1, na.rm = T),
            arrDelay_rate = n_ArrDelay/totalFlights,
            
            n_OnTime_DEP = totalFlights - n_DepDelay,
            onTimeDep_pct = n_OnTime_DEP/totalFlights,
            
            n_OnTime_ARR = totalFlights - n_ArrDelay,
            onTimeArr_pct = n_OnTime_ARR/totalFlights,
            
            n_CANCELLED = sum(CANCELLED == 1, na.rm = T),
            cancel_rate = n_CANCELLED/totalFlights,
            
            # Time structure
            sd_dep_time = sd(DEP_TIME, na.rm = TRUE),
            peak_dep_Count = sum(DEP_TIME >= 0700 & 
                          DEP_TIME <= 1900,na.rm = TRUE),
            peak_dep_share = mean(DEP_TIME >= 700 & 
                      DEP_TIME <= 1900,na.rm = TRUE),
  
            avgDepDelay_time = mean(DEP_DELAY_NEW, na.rm = T),
            avgArrDelay_time = mean(ARR_DELAY_NEW, na.rm = T),
            .groups = "drop") |>
  
  # route performance
  group_by(YEAR, MONTH, ORIGIN, DEST) |>
  mutate(route_total_flights = sum(totalFlights)) |>
  ungroup() |>

  # airline performance from route average
  mutate(
    airlineRouteShare = totalFlights/route_total_flights) 
  
  hub_Share_df <- clean |>
    group_by(YEAR, MONTH, BRANDED_CODE_SHARE, ORIGIN) |>
    summarise(num_carrier_origin_flights = n(), .groups = "drop") |>
    group_by(ORIGIN, YEAR, MONTH) |>
    mutate(monthly_departing_flights = sum(num_carrier_origin_flights)) |>
    ungroup() |>
    mutate(hubShare_Month = num_carrier_origin_flights/monthly_departing_flights)
    
    routeLevel_df <- routeLevel_df |> left_join(hub_Share_df, by = c("YEAR", "MONTH", "BRANDED_CODE_SHARE", "ORIGIN"))
    
    return(routeLevel_df)
}


all_files <- list.files(
  path = "/Users/kiepeeh1/Documents/Academic_Research/ECON_399_Industrial Organization/BTS Data/Flights Data",
  pattern = "20[2-5][0-9]\\.csv",
  full.names = TRUE
)

airline_route_level_data <- map_dfr(all_files, process_month, .id = "file_id")
```

```{r}
write_csv(airline_route_level_data, "~/Documents/Academic_Research/econometrics-airport-dominance-impact-on-service-quality/00_data/processed/airline_route_level_data.csv")
```

